Online Classification of arXiv Papers Based on Your Preferences

### 2014-05-16 Paper.tex
some edits to paper.

In Section 3, we describe the main components of the program in the python modules we created, and give a description on how to use the tool. We present our results in Section 4, followed by other observations and concluding remarks in seciont 5.

@@

\section{Theory: Algorithm and Bounding Updates}

The main engine behind the tool is a modified Perceptron algorithm which employs an accept-reject method among the papers it labels uninteresting. The following pseudocode details the algorithm, where $x_t$ denotes feature vectors, $y_t$ denotes labels, $\widehat{y}_t$ are predictions, $w_t$ are hypotheses, and $k_t$ are i.i.d. random variables used for the accept-reject method. $k_t = 1$ with probability $p$ and $k_t = 0$ with probability $1-p$. As mentioned, if the algorithm only made updates when $\widehat{y}=1$ (i.e. interesting) then this would leave the number of mistakes made when $\widehat{y} = -1$ unchecked (hence, several lost interesting articles). As a compromise, we generate the i.i.d. variables $k_t$ for each observation when $\widehat{y} = -1$. Note that if $\widehat{y} = -1$ and $k=1$, the algorithm will ask for the correct label and make an update. Otherwise if $\widehat{y} = -1$ and $k=0$, the algorithm does not ask for the correct label and makes no update.\\

@@

Let $I_1$ be the set of $t$ where a mistake was made and
$\widehat{y}=1$ (false positives).  Let $I_2$ be the set where a mistake was made with
$\widehat{y} = -1$ and $k = 1$ (false negatives where the algorithm updates), and $I_3$ the set where a mistake was
made with $\widehat{y} = -1$ and $k = 0$ (false negatives where the algorithm does {\it not} make an update). Finally, let $M_i = |I_i|$ for $i =
1,2,3$. We bound $M_1 + M_2$ in the same way we bound $M$ in the original Perceptron algorithm. Assume there exists a $\rho$ and $R$ satisfying the assumptions of Theorem 7.8 from \cite{FML}, and a separating hyperplane determined by the vector $v$. Let $M' = M_1 + M_2$, then
\begin{align*}
M' \rho =&\ \frac{v}{\|v\|}\sum_{I_1 \cup I_2} y_tx_t \
 \leq   \Bigl\|\sum_{t \in I_1 \cup I_2} y_t x_t\Bigr\| \\
=& \ \Bigl\|\sum_{t \in I_1 \cup I_2} (w_{t+1} - w_t) \Bigr\| 
=  \|w_{T+1} \| \\
=&\ \sqrt{\sum_{t \in I_1 \cup I_2} \|  w_{t+1}\|^2 - \|w_t\|^2 }\\
= &\ \sqrt{\sum_{t \in I_1 \cup I_2} \|  w_{t} + y_t x_t\|^2 - \|w_t\|^2 }\\
= &\ \sqrt{\sum_{t \in I_1 \cup I_2} 2y_t w_t\cdot x_t - \|x_t\|^2} \\
\leq &\ \sqrt{\sum_{t \in I_1 \cup I_2} \|x_t\|^2}
\leq \sqrt{M'r^2}.
\end{align*}
This implies that $M' \leq \frac{r^2}{\rho^2}$. Since $M = M' + M_3$, a bound on $M_3$ is needed. We revert to a weaker bound and consider the expectation of
$M_3$. Since 
\[M_3 = \sum_{t=1}^T 1_{\left\{\widehat{y}_t = -1 \cap y_t = 1\right\}}1_{k_t = 0}, \]
we get that
\begin{align*}
\E[M_3] = & \sum_{t=1}^T \E[1_{\left\{\widehat{y_t} = -1 \cap y_t = 1\right\}}]\E[1_{k_t = 0}] \\
 = &\ (1 - p)\sum_{t=1}^T \E[1_{\left\{\widehat{y_t} = -1 \cap y_t = 1\right\}}],
\end{align*}
where the last equality uses the independence of $k_t$ from the data. To control the sum of expectations in this expression we consider the expectation of $M_2$. Since
\[
M_2 =  \sum_{t=1}^T 1_{\left\{\hat{y_t} = -1 \cap y_t = 1\right\}}1_{k_t = 1},
\]
then
\begin{align*}
\E[M_2] =&\ \sum_{t=1}^T \E[1_{\left\{\hat{y_t} = -1 \cap y_t = 1\right\}}]\E[1_{k_t = 1}] \\
=&\ p\sum_{t=1}^T \E[1_{\left\{\hat{y_t} = -1 \cap y_t = 1\right\}}].
\end{align*}
Thus
\[
E[M_3] = \ \frac{1-p}{p}\ \E[M_2].
\]
We can apply the crude bound $\E[M_3] \leq \frac{1-p}{p}\ \E[M']$, since $M_2 \leq M'$. Thus the expected number of mistakes can be bounded by
\begin{align*}
\E[M]  =  &\ \E[M'] + \E[M_3] \\
\leq & \ \frac{r^2}{\rho^2} + \frac{1-p}{p}\frac{r^2}{\rho^2} \\
\leq  & \ \frac{1}{p}\frac{r^2}{\rho^2}
\end{align*}
Note that this is a weak bound.  Not only is it greater than the
original perceptron bound by a factor of $\frac{1}{p}$, it is also only a
bound in expectation.

@@

There are some additional Python files included that we used primarily for testing. For instance, {\tt getTestData.py} and {\tt LabelData.py} are scripts that call the functions that collect arXiv data and present the abstracts to acquire labels, respectively, which we place in a single script for the user. Additionally {\tt ClassifierTester.py} defines a class for the classification algorithms we include allowing them to be used in a uniform manner despite taking different parameters. This enables us to run tests over each algorithm, with multiple data sets from a few different users, using a set of several values for each parameter in a fairly organized manner in the script {\tt TestClassifiers.py}. Plots are also constructed in that script as well.

To run the tool as a user, we've included a script {\tt Arxiver.py}, which first asks for filters to draw articles including subject, subfields, and dates. Then it presents a subset of those it pulled from arXiv, filtering out some of those it predicts are uninteresting according to the Perceptron algorithm we've described. When finished {\tt Arxiver} prints out the arXiv ID's for those articles the user labels as interesting.

An important aspect of any text classification device is the data used to construct feature vectors. There are several approaches to text classification, itself a large field of research. We take an elementary approach and use a bag-of-words model; that is, performing a word count for each distinct word in the abstract and converting that into a vector. 
Each distinct word corresponds to a different coordinate in the feature space. This ``vectorization" is defined in a class {\tt TextFeatureVector} in {\tt ArxivData.py} and is called in the function {\tt FeaturizeAbstracts()}, which appears before running the algorithms, naturally. Word counts seems like a naive approach, yet the bag-of-words model is well used and often augmented with other methods. Moreover implementing such a model is not terribly complicated, which makes it a useful and attractive option.

@@

It was interesting to see how varying $p$ the probability of presenting the paper had an impact on the results, some which might not have been entirely expected. For instance I would expect that asking for the most labels from the user (i.e. $p = 1$) would translate to considerably higher gains in both precision and recall all around. Yet for the best performing algorithms there wasn't a huge difference in recall between the adaptive version and $p = 1$.  For precision, the lower $p$ algorithms actually performed comparably to the original schemes.

\begin{figure}\centering
\includegraphics[width=0.7\textwidth]{EveryoneRecallPaper.pdf}
\includegraphics[width=0.7\textwidth]{EveryonePrecisionPaper.pdf}
\label{fig:PrecisionandRecall}
\caption{Performance of tested online classification algorithms using different values of $p$.  The adaptive $p$ algorithm performs relatively well compared to the original algorithm ($p = 1$), and requires less than half as many negative updates. The $p = 0.1$ case tends to perform poorly, but requires far fewer updates.  The majority vote classifier combines all others in this figure.}
\end{figure}

@@

In order for this approach to be useful, it cannot present the user with every single prediction.  Instead, we weaken the algorithms to present the user with all articles it predicts will be interesting, and only some of those that it predicts will not be interesting.  We show that in this context, a bound on the expected number of mistakes can still be achieved under the same assumptions as Theorem 7.8 from \cite{FML}.  Finally, we test several algorithms on labeled data, and show that they perform comparably to the original learning algorithms, but require less input from the user.  \\
In the future, it would be nice to develop tighter theoretical bounds on our modified learning algorithm, and particularly to develop an improved bound on the adaptive $p$ algorithm.  It is possible that such a bound would lead to a more informed decision on how to change $p$ in the algorithm to minimize mistakes. Perhaps giving a different functional form to the way $p$ is changed can produce an algorithm that minimizes the number of labels required from the user.\\
Finally, it would be worthwhile to explore other online learning algorithms and possibly other features that might improve predictive performance. Considering a wider variety of text classification models would probably produce better results, and the limited time to develop this tool compromised our ability to develop and play around with some of these models.
\bibliography{paper}
\end{document}

>>>>

In Section 3, we describe the main components of the program in the python modules we created, and give a description on how to use the tool. We present our results in Section 4, followed by other observations and concluding remarks in Section 5.

@@

\section{Theory: Algorithm and Bounding Updates}
\label{sec:theory}
The main engine behind the tool is a modified Perceptron algorithm which employs an accept-reject method among the papers it labels uninteresting. The following pseudocode details the algorithm, where $x_t$ denotes feature vectors, $y_t$ denotes labels, $\widehat{y}_t$ are predictions, $w_t$ are weights, and $k_t$ are i.i.d. random variables used for the accept-reject method. $k_t = 1$ with probability $p$ and $k_t = 0$ with probability $1-p$. As mentioned, if the algorithm only made updates when $\widehat{y}=1$ (i.e. interesting) then this would leave the number of mistakes made when $\widehat{y} = -1$ unchecked (hence, several lost interesting articles). As a compromise, we generate the i.i.d. variables $k_t$ for each observation when $\widehat{y} = -1$. If $\widehat{y} = -1$ and $k=1$, the algorithm will ask for the correct label and make an update. Otherwise if $\widehat{y} = -1$ and $k=0$, the algorithm does not ask for the correct label and makes no update.\\

@@

Let $I_1$ be the set of $t$ where a mistake was made and
$\widehat{y}=1$ (false positives).  Let $I_2$ be the set where a mistake was made with
$\widehat{y} = -1$ and $k = 1$ (false negatives where the algorithm updates), and $I_3$ the set where a mistake was
made with $\widehat{y} = -1$ and $k = 0$ (false negatives where the algorithm does {\it not} make an update). Finally, let $M_i = |I_i|$ for $i =
1,2,3$. We bound $M_1 + M_2$ in the same way we bound $M$ in the original Perceptron algorithm, because these are all the mistakes after which an update is made.. Assume there exists a $\rho$ and $R$ satisfying the assumptions of Theorem 7.8 from \cite{FML}, and a separating hyperplane determined by the vector $v$. Let $M' = M_1 + M_2$, then
\begin{align*}
M' \rho =&\ \frac{v}{\|v\|}\sum_{I_1 \cup I_2} y_tx_t \
 \leq   \Bigl\|\sum_{t \in I_1 \cup I_2} y_t x_t\Bigr\| \\
=& \ \Bigl\|\sum_{t \in I_1 \cup I_2} (w_{t+1} - w_t) \Bigr\| 
=  \|w_{T+1} \| \\
=&\ \sqrt{\sum_{t \in I_1 \cup I_2} \|  w_{t+1}\|^2 - \|w_t\|^2 }\\
= &\ \sqrt{\sum_{t \in I_1 \cup I_2} \|  w_{t} + y_t x_t\|^2 - \|w_t\|^2 }\\
= &\ \sqrt{\sum_{t \in I_1 \cup I_2} 2y_t w_t\cdot x_t - \|x_t\|^2} \\
\leq &\ \sqrt{\sum_{t \in I_1 \cup I_2} \|x_t\|^2}
\leq \sqrt{M'r^2}.
\end{align*}
This implies that $M' \leq \frac{r^2}{\rho^2}$. Since $M = M' + M_3$, a bound on $M_3$ is needed. We revert to a weaker bound and consider the expectation of
$M_3$. Since 
\[M_3 = \sum_{t=1}^T 1_{\left\{\widehat{y}_t = -1 \cap y_t = 1\right\}}1_{k_t = 0}, \]
we get that
\begin{align*}
\E[M_3] = & \sum_{t=1}^T \E[1_{\left\{\widehat{y_t} = -1 \cap y_t = 1\right\}}]\E[1_{k_t = 0}] \\
 = &\ (1 - p)\sum_{t=1}^T \E[1_{\left\{\widehat{y_t} = -1 \cap y_t = 1\right\}}],
\end{align*}
where the last equality uses the independence of $k_t$ from the data, and the expecation is taken over the realizations of $k_t$. To control the sum of expectations in this expression we consider the expectation of $M_2$. Since
\[
M_2 =  \sum_{t=1}^T 1_{\left\{\hat{y_t} = -1 \cap y_t = 1\right\}}1_{k_t = 1},
\]
then
\begin{align*}
\E[M_2] =&\ \sum_{t=1}^T \E[1_{\left\{\hat{y_t} = -1 \cap y_t = 1\right\}}]\E[1_{k_t = 1}] \\
=&\ p\sum_{t=1}^T \E[1_{\left\{\hat{y_t} = -1 \cap y_t = 1\right\}}].
\end{align*}
Thus
\[
E[M_3] = \ \frac{1-p}{p}\ \E[M_2].
\]
We can apply the crude bound $\E[M_3] \leq \frac{1-p}{p}\ \E[M']$, since $M_2 \leq M'$. Thus the expected number of mistakes can be bounded by
\begin{align*}
\E[M]  =  &\ \E[M'] + \E[M_3] \\
\leq & \ \frac{r^2}{\rho^2} + \frac{1-p}{p}\frac{r^2}{\rho^2} \\
\leq  & \ \frac{1}{p}\frac{r^2}{\rho^2}
\end{align*}
Note that this is a weak bound.  Not only is it greater than the
original perceptron bound by a factor of $\frac{1}{p}$, it is also only a
bound in expectation.

@@

There are some additional Python files included that we used primarily for testing. For instance, {\tt getTestData.py} and {\tt LabelData.py} are scripts that call the functions that collect arXiv data and present the abstracts to acquire labels, respectively, which we place in a single script for the user. Additionally {\tt ClassifierTester.py} takes a single classifier and tests it on a grid of possible parameters to find the best values. This enables us to run tests over each algorithm, with multiple data sets from a few different users, using a set of several values for each parameter in a fairly organized manner in the script {\tt TestClassifiers.py}. Plots are also constructed in that script as well.\\

To run the tool as a user, we've included a script {\tt Arxiver.py}, which first asks for filters to draw articles including subjects and subcategories. Then it presents abstracts from a subset of those it pulled from arXiv.  As described in Section \ref{sec:theory}, it only asks for some of labels from papers that it predicts are uninteresting. As it runs, {\tt Arxiver} automatically saves the articles that the user labels as interesting.\\

An important aspect of any text classification device is the data used to construct feature vectors. There are several approaches to text classification, itself a large field of research. We take an elementary approach and use a bag-of-words model; that is, performing a word count for each distinct word in the abstract and converting that into a vector.  We normalize this vector by the total length of the abstract,  
Each distinct word corresponds to a different coordinate in the feature space. This ``vectorization" is defined in a class {\tt TextFeatureVector} in {\tt ArxivData.py} and TextFeatureVector objects are created from abstracts using {\tt FeaturizeAbstracts()}. Word counts seems like a naive approach, yet the bag-of-words model is well used and often augmented with other methods. Moreover implementing such a model is not terribly complicated, which makes it a useful and attractive option.   \\
All of the software used for this tool project is available at ``github.com/skwaap/fml-arxiv'' including scripts for testing.\\

@@

It was interesting to see how varying $p$ the probability of presenting the paper had an impact on the results.  For the best performing algorithms there wasn't a huge difference in recall between the adaptive version and $p = 1$.  For precision, the lower $p$ algorithms actually performed comparably to the original schemes.  This is not entirely unexpected, since precision measures the number of false positives, and the algorithms outlined here update on all false positives independent of $p$. \\

\begin{figure}[h]\centering
\includegraphics[width=0.65\textwidth]{EveryoneRecallPaper.pdf}
\includegraphics[width=0.65\textwidth]{EveryonePrecisionPaper.pdf}
\label{fig:PrecisionandRecall}
\caption{Performance of tested online classification algorithms using different values of $p$.  The adaptive $p$ algorithm performs relatively well compared to the original algorithm ($p = 1$), and requires less than half as many negative updates. The $p = 0.1$ approach has ver poor recall, but requires far fewer updates.  The majority vote classifier combines all others in this figure.}
\end{figure}

@@

In order for this approach to be useful, it cannot present the user with every single prediction.  Instead, we weaken the algorithms to present the user with all articles it predicts will be interesting, and only some of those that it predicts will not be interesting.  We show that in this context, a bound on the expected number of mistakes can still be achieved under the same assumptions as Theorem 7.8 from \cite{FML}.  Finally, we test several algorithms on labeled data, and show that they perform comparably to the original learning algorithms, but require less input from the user.  \\
In the future, it would be nice to develop tighter theoretical bounds on our modified learning algorithm, and particularly to develop an improved bound on the adaptive $p$ algorithm.  It is possible that such a bound would lead to a more informed decision on how to change $p$ in the algorithm to minimize mistakes. Perhaps giving a different functional form to the way $p$ is changed can produce an algorithm that minimizes the number of labels required from the user, and potentially give a clear improvement over the fixed $p$ case.\\
Finally, it would be worthwhile to explore other online learning algorithms and possibly other features that might improve predictive performance. Considering a wider variety of text classification models would probably produce better results, and the limited time to develop this tool compromised our ability to develop and play around with some of these models.
\bibliography{paper}
\end{document}
